NONLINEAR OBSERVABILITY FOR RELATIVE SATELLITE ORBITS WITH ANGLES-ONLY MEASUREMENTS

### 2018-06-09 Conference/AAS 2015 OD EKF/ISSFD14/ISSFD14.tex
added Conference folder

% use paper, or submit
% use 11 pt (preferred), 12 pt, or 10 pt only

%\documentclass[letterpaper, preprint, paper,11pt]{AAS}	% for preprint proceedings
%\documentclass[letterpaper, paper,11pt]{AAS}		% for final proceedings (20-page limit)
%\documentclass[letterpaper, paper,12pt]{AAS}		% for final proceedings (20-page limit)
\documentclass[letterpaper, paper,10pt]{AAS}		% for final proceedings (20-page limit)
%\documentclass[letterpaper, submit]{AAS}			% to submit to JAS

\usepackage{bm}
\usepackage{amssymb,amsmath,amsthm,times,graphicx,subfigure,tabularx,booktabs,colortbl,multirow,threeparttable}
\usepackage{subfigure}
%\usepackage[notref,notcite]{showkeys}  % use this to temporarily show labels
\usepackage[colorlinks=true, pdfstartview=FitV, linkcolor=black, citecolor= black, urlcolor= black]{hyperref}
\usepackage{overcite}
\usepackage{footnpag}			      	% make footnote symbols restart on each page

\usepackage{color}
\newcommand{\hilight}[1]{\colorbox{green}{#1}}

\newtheorem{prop}{Proposition}

\newcommand{\abs}{\mathop{\mathrm{abs}}\nolimits}
\newcommand{\diag}{\mathop{\mathrm{diag}}\nolimits}
\newcommand{\norm}[1]{\ensuremath{\left\| #1 \right\|}}
\newcommand{\bracket}[1]{\ensuremath{\left[ #1 \right]}}
\newcommand{\braces}[1]{\ensuremath{\left\{ #1 \right\}}}
\newcommand{\parenth}[1]{\ensuremath{\left( #1 \right)}}
\newcommand{\pair}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\met}[1]{\ensuremath{\langle\langle #1 \rangle\rangle}}
\newcommand{\refeqn}[1]{(\ref{eqn:#1})}
\newcommand{\reffig}[1]{Fig. \ref{fig:#1}}
\newcommand{\tr}[1]{\mathrm{tr}\ensuremath{\negthickspace\bracket{#1}}}
\newcommand{\trs}[1]{\mathrm{tr}\ensuremath{[#1]}}
\newcommand{\ave}[1]{\mathrm{E}\ensuremath{[#1]}}
\newcommand{\deriv}[2]{\ensuremath{\frac{\partial #1}{\partial #2}}}
\newcommand{\SO}{\ensuremath{\mathsf{SO(3)}}}
\newcommand{\T}{\ensuremath{\mathsf{T}}}
\renewcommand{\L}{\ensuremath{\mathsf{L}}}
\newcommand{\so}{\ensuremath{\mathfrak{so}(3)}}
\newcommand{\SE}{\ensuremath{\mathsf{SE(3)}}}
\newcommand{\se}{\ensuremath{\mathfrak{se}(3)}}
\renewcommand{\Re}{\ensuremath{\mathbb{R}}}
\newcommand{\aSE}[2]{\ensuremath{\begin{bmatrix}#1&#2\\0&1\end{bmatrix}}}
\newcommand{\ase}[2]{\ensuremath{\begin{bmatrix}#1&#2\\0&0\end{bmatrix}}}
\newcommand{\D}{\ensuremath{\mathbf{D}}}
\renewcommand{\d}{\ensuremath{\mathfrak{d}}}
\newcommand{\Sph}{\ensuremath{\mathsf{S}}}
\renewcommand{\S}{\Sph}
\newcommand{\J}{\ensuremath{\mathbf{J}}}
\newcommand{\Ad}{\ensuremath{\mathrm{Ad}}}
\newcommand{\intp}{\ensuremath{\mathbf{i}}}
\newcommand{\extd}{\ensuremath{\mathbf{d}}}
\newcommand{\hor}{\ensuremath{\mathrm{hor}}}
\newcommand{\ver}{\ensuremath{\mathrm{ver}}}
\newcommand{\dyn}{\ensuremath{\mathrm{dyn}}}
\newcommand{\geo}{\ensuremath{\mathrm{geo}}}
\newcommand{\Q}{\ensuremath{\mathsf{Q}}}
\newcommand{\G}{\ensuremath{\mathsf{G}}}
\newcommand{\g}{\ensuremath{\mathfrak{g}}}
\newcommand{\Hess}{\ensuremath{\mathrm{Hess}}}

\newcommand{\x}{\ensuremath{\mathbf{x}}}
\renewcommand{\r}{\mathbf{r}}
\renewcommand{\u}{\mathbf{u}}
\newcommand{\y}{\mathbf{y}}

\newcommand{\bfi}{\bfseries\itshape\selectfont}

%\renewcommand{\baselinestretch}{1.2}

\PaperNumber{23-451}

\begin{document}

\title{Nonlinear Observability Measure for Relative Orbit Determination with Angles-Only Measurements}

\author{Evan Kaufman\thanks{Doctoral Student, Mechanical and Aerospace Engineering, George Washington University, 801 22nd St NW, Washington, DC 20052, Email: evankaufman@gwu.edu.},  
T. Alan Lovell\thanks{Research Aerospace Engineer, Air Force Research Laboratory, Space Vehicles Directorate, Kirtland AFB, NM.},
\ and Taeyoung Lee\thanks{Assistant Professor, Mechanical and Aerospace Engineering, George Washington University, 801 22nd St NW, Washington, DC 20052, Tel: 202-994-8710, Email: tylee@gwu.edu.}
\thanks{This research has been supported in part by NSF under the grants CMMI-1243000 (transferred from 1029551), CMMI-1335008, and CNS-1337722.}
}

\maketitle{}

\begin{abstract}
%An angles-only extended Kalman filter (EKF), based off of two-body motion, is developed to estimate the position of a Deputy satellite relative to a chief satellite.
A new nonlinear observability measure is proposed for relative orbit determination when lines-of-sight between satellites are measured only. It corresponds to a generalization of the observability Gramian in linear dynamic systems to the nonlinear relative orbit dynamics represented by the two-body problems. An extended Kalman filter (EKF) is adapted to this problem and is evaluated with various gravitational harmonics and initial orbital determination (IOD) predictions. Extensive results illustrate correspondence between the proposed observability measure with filtering errors. An extensive numerical analysis in realistic scenarios includes satellite propagation of the two-body problem the $J_2$ perturbation effects.
\end{abstract}

\section{Introduction}

Space-based surveillance or relative navigation is desirable for many spacecraft missions, such as formation control and rendezvous. Spacecraft maneuvers based only on on-board measurements reduce the total operating cost significantly, and it improves safety against communication interruptions with ground stations.  Relative navigation between spacecraft in close-proximity essentially corresponds to space-based orbit determination.  In particular, vision-based navigation and estimation of relative orbits have received attention recently, since optical sensors have the desirable properties of low cost and minimal maintenance, while providing accurate line-of-sight measurements, or angles from a chief to a deputy.

@@

All of these results are based on linear relative orbital dynamics. It is straightforward to see that the relative orbit is not observable with angles-only measurements through its linearized dynamics. In other words, there are an infinite number of relative orbits that yield the identical line-of-sight measurements, and the orbital distance between a deputy and a chief cannot be determined by angles-only measurements. As such, it is required to study the nonlinear relative orbital dynamics to determine observability with angles-only measurements. Linear observability analysis is performed numerically for a particular case in~\cite{YimCraPASMM04}.

Recently, observability criteria have been derived for the nonlinear relative orbital dynamics represented by the solutions of the two-body problem without linearization~\cite{LovLeePISSFD14}.  Assuming that a chief is on a circular orbit with a prescribed orbital radius, nonlinear equations of motion for the relative orbital motion of a deputy with respect to the chief are presented.  A differential geometric method, based on the Lie derivatives of the line-of-sight from the chief to the deputy, is applied to obtain sufficient conditions for observability.  It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements.

However, this result does not provide any information of the degree of observability. The main contribution of this paper is proposing a new  measure of observability for relative orbit determination with angles-only measurements. For a given initial condition of relative orbits, the proposed observability measure determines how much the corresponding relative orbits are easy or difficult to estimate, thereby providing a quantitative measure for relative orbit determination.

Furthermore, this paper serves as a feasibility study of an extended Kalman filter (EKF) for orbit determination using angles-only measurements in realistic scenarios. The filter is subject to measurement uncertainty and perturbation forces not considered with the two-body problem, on which the filter is based.
%The EKF is simulated with various time steps to show how the filter performs on a wide range of systems with different sample rates.
Then, the performance of the EKF is compared with the proposed observability measure to show the correspondence.

@@

Consider two satellites orbiting around the Earth, where each satellite is modeled as a rigid body. Suppose that a \textit{chief} satellite is on a circular orbit with a pre-determined orbital radius of $a\in\Re$.

@@

Nonlinear equations of motion for the relative motion of the deputy with respect to the chief can be derived as follows based on Lagrangian mechanics~\cite{LovLeePISSFD14}.

@@

We assume that the line-of-sight from the chief to the deputy is measured by onboard sensing, such as optical sensors. The measurement is represented by the unit-vector of the relative position vector, i.e.,
\begin{align}
\y = h(\mathrm{x}) =\frac{\r}{\|\r\|},\label{eqn:y}
\end{align}
where $\y\in\Re^M$ with $M=3$, and it satisfies $\|\y\|=1$.

@@

Based on the nonlinear dynamic model presented at the previous section, here we analyze the observability of the relative orbit with line-of-sight measurements.
This section serves as a summary of~\cite{LovLeePISSFD14}.

@@

Observability of nonlinear systems has been studied in~\cite{HerKreITAC77}, and it is summarized as follows. For a given nonlinear dynamic system described with \refeqn{xxdot} and \refeqn{y}, a pair of points $\x_0$ and $\x_1$ are called \textit{indistinguishable} if the outputs of the corresponding solutions starting from each of $\x_0$ and $\x_1$ are identical for a certain time period. The systems is \textit{locally weakly observable} at $\x_0$, if there exists an open neighborhood $V$ of $\x_0$ such that for every open neighborhood $U$ of $\x_0$ contained in $V$, the only indistinguishable point to $\x_0$ is the point $\x_0$ itself.

The Lie-derivative of the output $h(\x)$ along $f(\x)$ is defined as follows:
\begin{align*}
L_f h(\x) = \deriv{h(\x)}{\x}f(\x)\in\Re^{M\times 1},
\end{align*}
which corresponds to the directional derivative of $h(\x)$ along $f(\x)$. For a non-negative integer $i$, the $i$-th order Lie-derivative is defined by induction as $L_f^i h = L_f (L_f^{i-1} h)$ with $L_f^0 h = h$. Define an observability matrix $\mathcal{O}\in\Re^{NM\times N}$ as
\begin{align*}
\mathcal{O}(\x_0) = \deriv{}{\x} \begin{bmatrix} L_f^0 h(\x) \\ L_f^1 h(\x)\\ \vdots \\L_f^{N-1} h(\x)\end{bmatrix}\bigg|_{\x=\x_0}.
\end{align*}
It has been shown that the system is locally weakly observable at $\x_0$ if the rank of the observability matrix $\mathcal{O}(\x_0) = N$. When applied to linear dynamics, this yields the well-known observability rank condition for linear systems. Note that when there is more than a single measurement, i.e., $M>1$, the observability rank condition can be satisfied without a need for computing the higher-order Lie derivatives up to the $N-1$-th order.

@@

After straightforward but tedious algebraic manipulations using the following identity repeatedly,
\begin{align*}
\delta \parenth{\frac{1}{\|\mathbf{r}\|^i}} 
& = -i\frac{ \mathbf{r}^T \delta \mathbf{r}}{\|\mathbf{r}\|^{i+2}},
\end{align*}
for any positive integer $i$, we can show that each of the sub-matrices $\mathcal{O}_{ij}$ of the observability matrix $\mathcal{O}$ is given by
\begin{align}
\mathcal{O}_{00} & = \deriv{\y}{\r} = \frac{1}{\|\r\|}(I - \y\y^T),\label{eqn:O_00}\\
\mathcal{O}_{10} & = \deriv{\dot\y}{\r} = \deriv{}{\r} \parenth{\deriv{\y}{\r}\dot \r}= -\frac{1}{\|\r\|^2}\braces{
{\dot\r\y^T +\y^T\dot\r I +\y\dot\r^T}
- 3\y\y^T\dot\r \y^T},\\
\mathcal{O}_{11} & = \deriv{\dot\y}{\dot\r} = \deriv{}{\dot\r} \parenth{\deriv{\y}{\r}\dot \r}= \deriv{\y}{\r} = \mathcal{O}_{00},\label{eqn:O11}\\
\mathcal{O}_{20} & = \deriv{\ddot\y}{\r} = -\dfrac{2\dot\r \dot\r^T  +\dot\r^T\dot\r I}{\|\r\|^3}
+3\dfrac{2(\r^T\dot\r) \dot\r\r^T +(\dot\r^T\dot\r)\r\r^T
+(\r^T\dot\r)^2 I
+2(\r^T\dot\r)\r  \dot\r^T
}{\|\r\|^5}
- 15 \dfrac{(\r^T\dot\r)^2\r  \r^T}{\|\r\|^7}\nonumber \\
&\quad -\dfrac{\ddot{\r}\r^T +\r^T\ddot{\r} I +\r \ddot{\r}^T}{\|\r\|^3}
+ 3 \dfrac{(\r^T\ddot{\r})\r \r^T}{\|\r\|^5}
+\parenth{\dfrac{I}{\|\mathbf{r}\|} - \dfrac{\mathbf{r}\mathbf{r}^T}{\|\mathbf{r}\|^3}}
\parenth{-[\omega]_\times^2 -\dfrac{\mu I_{3\times 3}}{\|\r_a\|^3} + \dfrac{3\mu\r_a\r_a^T}{\|\r_a\|^5}},\\
\mathcal{O}_{21} & = \deriv{\ddot\y}{\dot\r} = \deriv{}{\dot\r}\parenth{\deriv{}{\r}\parenth{\deriv{\y}{\r}\dot\r}\dot\r + \deriv{\y}{\r}\ddot\r} = 2 \deriv{}{\r} \parenth{\deriv{\y}{\r}\dot\r} + \deriv{\y}{\r} \deriv{\ddot\r}{\dot\r}
\nonumber\\
& \quad = 2\mathcal{O}_{10} 
- 2 \mathcal{O}_{00}[\omega]_\times,\label{eqn:O_21}
\end{align}
where $[\omega]_\times\in\Re^{3\times 3}$ is defined as
\begin{align*}
[\omega]_\times = \begin{bmatrix} 0 & -n & 0 \\ n & 0 & 0 \\ 0 & 0 & 0 \end{bmatrix}.
\end{align*}
These expressions were fully verified by the Matlab symbolic computation toolbox.

@@

\begin{prop}
Define three vectors $\mathbf{v}_{rel}$, $\mathbf{a}_1,\mathbf{a}_2\in\Re^3$ as 
\begin{align}
\mathbf{v}_{rel} & = \dot\r +\omega\times\r,\label{eqn:vrel}\\
\mathbf{a}_1 & = \ddot\r-\deriv{\ddot\r}{\r}\r = -2\omega\times \dot\r - [\omega]_\times^2 a \mathbf{e}_1 -\dfrac{\mu a }{\|\r_a\|^3}\mathbf{e}_1- \dfrac{3\mu\r_a^T\r}{\|\r_a\|^5}\r_a,\label{eqn:a1}\\
\mathbf{a}_2 & = \ddot\r-\deriv{\ddot\r}{\r}\r -\deriv{\ddot\r}{\dot\r}\dot\r = \mathbf{a}_1+2\omega\times\dot\r= - [\omega]_\times^2 a \mathbf{e}_1 -\dfrac{\mu a }{\|\r_a\|^3}\mathbf{e}_1- \dfrac{3\mu\r_a^T\r}{\|\r_a\|^5}\r_a.\label{eqn:a2}
\end{align}
The nonlinear relative orbital dynamics are locally weakly observable at $\x=[\r^T,\dot\r^T]$ if
%\begin{list}{}{\setlength{\leftmargin}{1cm}\setlength{\itemsep}{0mm}\setlength{\parsep}{0mm}\setlength{\topsep}{0mm}\setlength{\parskip}{0mm}\setlength{\labelwidth}{2cm}}
%\item[(i)] when $\r\times\dot\r =0$, $\r\times\mathbf{a}_1\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_1)\neq 0$,
%\item[(ii)] when $\r\times\dot\r \neq 0$, $\r\times\mathbf{a}_2\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_2)\neq 0$.
%\end{list}
\begin{alignat}{2}
(i)&\; &\text{when $\r\times\dot\r =0$, $\r\times\mathbf{v}_{ref}\neq0$, $\r\times\mathbf{a}_1\neq 0$,  and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_1)\neq 0$},\label{eqn:cond1}\\
(ii)& &\text{when $\r\times\dot\r \neq 0$, $\r\times\mathbf{v}_{ref}\neq0$, $\r\times\mathbf{a}_2\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_2)\neq 0$.}\label{eqn:cond2}
\end{alignat}
\end{prop}

@@

We wish to show that there is no non-zero vector $\mathbf{c}$ satisfying \refeqn{Oc}. At the first three rows of \refeqn{Oc}, we have
\begin{align*}
\mathcal{O}_{00} \mathbf{c}_1 = \frac{1}{\|\r\|}(I-\y\y^T) \mathbf{c}_1 = 0.
\end{align*}
The matrix $\mathcal{O}_{00}$ has a one-dimensional null space spanned by $\y$. Therefore, without loss of generality, we can choose $\mathbf{c}_1=0$ or $\mathbf{c}_1=\r$. Suppose $\mathbf{c}_1=\r$ for the subsequent development.

@@

The third condition of \refeqn{cond1} implies that three vectors $\r$, $\mathbf{v}_{rel}$ and $\mathbf{a}_1$ do not belong to the same plane. The third condition of \refeqn{cond2} has a similar structure as well. This condition can be easily violated if the relative motion is planar, i.e., when $z(t)\equiv 0$ for all $t$.

>>>>

\documentclass[12pt,letterpaper]{ISSFD_v01}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amssymb,amsmath,amsthm,times,graphicx,subfigure,tabularx,booktabs,colortbl}

\usepackage{times} % use times font
\usepackage{mathptmx} %use same font for math mode
\DeclareSymbolFont{largesymbols}{OMX}{cmex}{m}{n} %undo the definition of large symbols from the mathptmx package
\usepackage{authblk} % use authblk package for author formatting
\renewcommand\Authfont{\bfseries} % make authors bold
\renewcommand\Affilfont{\normalfont\itshape} % make affiliation italic
\setlength{\affilsep}{0em} % no spacing between authors and affiliations
\usepackage[pdffitwindow=false,pdfstartview={FitH},colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black]{hyperref}
\usepackage{microtype} % makes detailed spacing look nicer.  This is only optional -- not required.

\usepackage{xspace}
\newcommand{\twentyfourth}{24\textsuperscript{th}\xspace}
\newcommand{\figText}{Fig.\ }
\newcommand{\tabText}{Tab.\ }
\newcommand{\eqText}{Eq.\ }
\newcommand{\norm}[1]{\ensuremath{\left\| #1 \right\|}}
\newcommand{\bracket}[1]{\ensuremath{\left[ #1 \right]}}
\newcommand{\braces}[1]{\ensuremath{\left\{ #1 \right\}}}
\newcommand{\parenth}[1]{\ensuremath{\left( #1 \right)}}
\newcommand{\pair}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\met}[1]{\ensuremath{\langle\langle #1 \rangle\rangle}}
\newcommand{\refeqn}[1]{(\ref{eqn:#1})}
\newcommand{\reffig}[1]{Fig. \ref{fig:#1}}
\newcommand{\tr}[1]{\mathrm{tr}\ensuremath{\negthickspace\bracket{#1}}}
\newcommand{\trs}[1]{\mathrm{tr}\ensuremath{[#1]}}
\newcommand{\ave}[1]{\mathrm{E}\ensuremath{[#1]}}
\newcommand{\deriv}[2]{\ensuremath{\frac{\partial #1}{\partial #2}}}
\newcommand{\SO}{\ensuremath{\mathsf{SO(3)}}}
\newcommand{\T}{\ensuremath{\mathsf{T}}}
\renewcommand{\L}{\ensuremath{\mathsf{L}}}
\newcommand{\so}{\ensuremath{\mathfrak{so}(3)}}
\newcommand{\SE}{\ensuremath{\mathsf{SE(3)}}}
\newcommand{\se}{\ensuremath{\mathfrak{se}(3)}}
\renewcommand{\Re}{\ensuremath{\mathbb{R}}}
\newcommand{\aSE}[2]{\ensuremath{\begin{bmatrix}#1&#2\\0&1\end{bmatrix}}}
\newcommand{\ase}[2]{\ensuremath{\begin{bmatrix}#1&#2\\0&0\end{bmatrix}}}
\newcommand{\D}{\ensuremath{\mathbf{D}}}
\renewcommand{\d}{\ensuremath{\mathfrak{d}}}
\newcommand{\Sph}{\ensuremath{\mathsf{S}}}
\renewcommand{\S}{\Sph}
\newcommand{\J}{\ensuremath{\mathbf{J}}}
\newcommand{\Ad}{\ensuremath{\mathrm{Ad}}}
\newcommand{\intp}{\ensuremath{\mathbf{i}}}
\newcommand{\extd}{\ensuremath{\mathbf{d}}}
\newcommand{\hor}{\ensuremath{\mathrm{hor}}}
\newcommand{\ver}{\ensuremath{\mathrm{ver}}}
\newcommand{\dyn}{\ensuremath{\mathrm{dyn}}}
\newcommand{\geo}{\ensuremath{\mathrm{geo}}}
\newcommand{\Q}{\ensuremath{\mathsf{Q}}}
\newcommand{\G}{\ensuremath{\mathsf{G}}}
\newcommand{\g}{\ensuremath{\mathfrak{g}}}
\newcommand{\Hess}{\ensuremath{\mathrm{Hess}}}

\renewcommand{\r}{\mathbf{r}}
\renewcommand{\u}{\mathbf{u}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\x}{\mathbf{x}}

\newcommand{\bfi}{\bfseries\itshape\selectfont}
\newtheorem{prop}{Proposition}

\title{NONLINEAR OBSERVABILITY FOR RELATIVE SATELLITE ORBITS
WITH ANGLES-ONLY MEASUREMENTS}

\author[(1)]{T. Alan Lovell}
\author[(2)]{Taeyoung Lee}
\affil[(1)]{Air Force Research Laboratory, 3550 Aberdeen Ave SE, Kirtland AFB, NM 87117 USA,\newline Tel: +1-505-853-4132, Email: splash73@yahoo.com}
\affil[(2)]{Department of Mechanical and Aerospace Engineering, The George Washington University,\newline
801 22nd St NW, Washington DC 20052, Tel: +1-202-994-8710, Email:
tylee@gwu.edu}

\date{}  % don't display a date

\begin{document}
\maketitle

\begin{abstract}
In this paper, nonlinear observability criteria are presented for the relative orbital dynamics represented by the solutions of the two-body problem. It is assumed that a chief is on a circular orbit with a prescribed orbital radius, and it measures line-of-sight toward a deputy. A differential geometric method, based on the Lie derivatives is used to derive sufficient conditions for observability of the orbital properties of deputy. It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements. An extended Kalman filter is also developed to numerically illustrate the observability of nonlinear relative orbits with angles-only measurements.
\end{abstract}

\begin{keywords}
Relative orbit, Observability, Line-of-sight measurement, Extended Kalman filter.
\end{keywords}

\section{Introduction}
Space-based surveillance or relative navigation is desirable for many spacecraft missions, such as formation control and rendezvous. Spacecraft maneuvers based only on on-board measurements reduce the total operating cost significantly, and it improves safety against communication interruptions with ground stations.  Relative navigation between spacecraft in close-proximity essentially corresponds to space-based orbit determination.  In particular, vision-based navigation and estimation of relative orbit have received attention recently, since optical sensors have desirable properties of low cost and minimal maintenance, while providing accurate line-of-sight measurements.%, or angles from a chief to a deputy.

@@

All of these results are based on linear relative orbital dynamics. It is straightforward to see that the relative orbit is not observable with angles-only measurements through its linearized dynamics, due the homogeneity property of linear dynamics implying that any solution of linear systems is directly proportional to its initial conditions. In other words, there are infinite number of relative orbits that yield the identical line-of-sight measurements, and the orbital distance between a deputy and a chief cannot be determined by angles-only measurements. As such, it is required to study the nonlinear relative orbital dynamics to determine observability with angles-only measurement. Linear observability analysis is performed numerically for a particular case in~\cite{YimCraPASMM04}. However, there have been no comprehensive analytic in nonlinear observability of relative orbits.

In this paper, the authors derive observability criteria for the nonlinear relative orbital dynamics represented by the solutions of the two-body problem without linearization.  Assuming that a chief is on a circular orbit with a prescribed orbital radius, nonlinear equations of motion for the relative orbital motion of a deputy with respect to the chief are presented.  A differential geometric method, based on the Lie derivatives of the line-of-sight from the chief to the deputy, is applied to obtain sufficient conditions for observability.  It is shown that under certain geometric conditions on the relative configuration between the chief and the deputy, the nonlinear relative motion is observable from angles-only measurements.  We also develop an extended Kalman filter to illustrate the observability properties numerically.

The main contribution of this paper is analytically confirming that the relative orbit is observable via line-of-sight measurements if its nonlinear dynamic characteristics are properly incorporated. To the authors' best knowledge, the sufficient conditions for nonlinear observability presented in this paper have not studied in prior publications.

This paper is organized as follows. The nonlinear dynamics of relative orbit is presented at Section 2, and observability criteria are developed at Section 3. These are followed by numerical examples of extended Kalman filtering and conclusions.

@@

Consider two satellites orbiting around the Earth, and each satellite is modeled as a rigid body. Suppose that a \textit{chief} satellite is on a circular orbit with a pre-determined orbital radius of $a\in\Re$.

@@

Nonlinear equations of motion for the relative motion of the deputy with respect to the chief can be derived as follows based on Lagrangian mechanics.

@@

We assume that the line-of-sight from the chief to the deputy is measured by an onboard sensor, such as optical sensors. The measurement is represented by the unit-vector of the relative position vector, i.e.,
\begin{align}
\y = h(\mathrm{x}) =\frac{\r}{\|\r\|},\label{eqn:y}
\end{align}
where $\y\in\Re^M$ with $M=3$, and it satisfies $\|\y\|=1$.

@@

Based on the nonlinear dynamic model presented at the previous section, here we present analyze the observability of the relative orbit with line-of-sight measurements.

@@

Observability of nonlinear systems has been studied in~\cite{HerKreITAC77}, and it is summarized as follows. For a given nonlinear dynamic system \refeqn{xxdot} and \refeqn{y}, a pair of points $\x_0$ and $\x_1$ are called \textit{indistinguishable} if the outputs of the corresponding solutions starting from each of $\x_0$ and $\x_1$ are identical for a certain time period. The systems is \textit{locally weakly observable} at $\x_0$, if there exists an open neighborhood $V$ of $\x_0$ such that for every open neighborhood $U$ of $\x_0$ contained in $V$, the only indistinguishable point to $\x_0$ is the point $\x_0$ itself.

Let the Lie-derivative of the output $h(\x)$ along $f(\x)$ as follows:
\begin{align*}
L_f h(\x) = \deriv{h(\x)}{\x}f(\x)\in\Re^{M\times 1},
\end{align*}
which corresponds to the directional derivative of $h(\x)$ along $f(\x)$. For a non-negative integer $i$, the $i$-th order Lie-derivative is defined by induction as $L_f^i h = L_f (L_f^{i-1} h)$ with $L_f^0 h = h$. Define an observability matrix $\mathcal{O}\in\Re^{NM\times N}$ as
\begin{align*}
\mathcal{O}(\x_0) = \deriv{}{\x} \begin{bmatrix} L_f^0 h(\x) \\ L_f^1 h(\x)\\ \vdots \\L_f^{N-1} h(\x)\end{bmatrix}\bigg|_{\x=\x_0}.
\end{align*}
It has been shown that the system is locally weakly observable at $\x_0$ if the rank of the observability matrix $\mathcal{O}(\x_0) = N$. When applied to linear dynamics, this yields the well-known observability rank condition for linear systems. Note that when there are more than a single measurement, i.e., $M>1$, the observability rank condition can be satisfied without need for computing the higher-order Lie derivatives up to the $N-1$-th order.

@@

After straightforward but tedious algebraic manipulations using the following identify repeatedly,
\begin{align*}
\delta \parenth{\frac{1}{\|\mathbf{r}\|^i}} 
& = -i\frac{ \mathbf{r}^T \delta \mathbf{r}}{\|\mathbf{r}\|^{i+2}},
\end{align*}
for any positive integer $i$, we can show that each of the sub-matrices $\mathcal{O}_{ij}$ of the observability matrix $\mathcal{O}$ is given by
\begin{align}
\mathcal{O}_{00} & = \deriv{\y}{\r} = \frac{1}{\|\r\|}(I - \y\y^T),\label{eqn:O_00}\\
\mathcal{O}_{10} & = \deriv{\dot\y}{\r} = \deriv{}{\r} \parenth{\deriv{\y}{\r}\dot \r}= -\frac{1}{\|\r\|^2}\braces{
{\dot\r\y^T +\y^T\dot\r I +\y\dot\r^T}
- 3\y\y^T\dot\r \y^T},\\
\mathcal{O}_{11} & = \deriv{\dot\y}{\dot\r} = \deriv{}{\dot\r} \parenth{\deriv{\y}{\r}\dot \r}= \deriv{\y}{\r} = \mathcal{O}_{00},\label{eqn:O11}\\
\mathcal{O}_{20} & = \deriv{\ddot\y}{\r} = -\dfrac{2\dot\r \dot\r^T  +\dot\r^T\dot\r I}{\|\r\|^3}
+3\dfrac{2(\r^T\dot\r) \dot\r\r^T +(\dot\r^T\dot\r)\r\r^T
+(\r^T\dot\r)^2 I
+2(\r^T\dot\r)\r  \dot\r^T
}{\|\r\|^5}
- 15 \dfrac{(\r^T\dot\r)^2\r  \r^T}{\|\r\|^7}\nonumber \\
&\quad -\dfrac{\ddot{\r}\r^T +\r^T\ddot{\r} I +\r \ddot{\r}^T}{\|\r\|^3}
+ 3 \dfrac{(\r^T\ddot{\r})\r \r^T}{\|\r\|^5}
+\parenth{\dfrac{I}{\|\mathbf{r}\|} - \dfrac{\mathbf{r}\mathbf{r}^T}{\|\mathbf{r}\|^3}}
\parenth{-[\omega]_\times^2 -\dfrac{\mu I_{3\times 3}}{\|\r_a\|^3} + \dfrac{3\mu\r_a\r_a^T}{\|\r_a\|^5}},\\
\mathcal{O}_{21} & = \deriv{\ddot\y}{\dot\r} = \deriv{}{\dot\r}\parenth{\deriv{}{\r}\parenth{\deriv{\y}{\r}\dot\r}\dot\r + \deriv{\y}{\r}\ddot\r} = 2 \deriv{}{\r} \parenth{\deriv{\y}{\r}\dot\r} + \deriv{\y}{\r} \deriv{\ddot\r}{\dot\r}
\nonumber\\
& \quad = 2\mathcal{O}_{10} 
- 2 \mathcal{O}_{00}[\omega]_\times,\label{eqn:O_21}
\end{align}
where $[\omega]_\times\in\Re^{3\times 3}$ is defined as
\begin{align*}
[\omega]_\times = \begin{bmatrix} 0 & -n & 0 \\ n & 0 & 0 \\ 0 & 0 & 0 \end{bmatrix}.
\end{align*}
These expressions were fully verified by the Matlab symbolic computation toolbox.

@@

\begin{prop}
Define three vectors $\mathbf{v}_{rel}$, $\mathbf{a}_1,\mathbf{a}_2\in\Re^3$ as 
\begin{align}
\mathbf{v}_{rel} & = \dot\r +\omega\times\r,\label{eqn:vrel}\\
\mathbf{a}_1 & = \ddot\r-\deriv{\ddot\r}{\r}\r = -2\omega\times \dot\r - [\omega]_\times^2 a \mathbf{e}_1 -\dfrac{\mu a }{\|\r_a\|^3}\mathbf{e}_1- \dfrac{3\mu\r_a^T\r}{\|\r_a\|^5}\r_a,\label{eqn:a1}\\
\mathbf{a}_2 & = \ddot\r-\deriv{\ddot\r}{\r}\r -\deriv{\ddot\r}{\dot\r}\dot\r = \mathbf{a}_1+2\omega\times\dot\r= - [\omega]_\times^2 a \mathbf{e}_1 -\dfrac{\mu a }{\|\r_a\|^3}\mathbf{e}_1- \dfrac{3\mu\r_a^T\r}{\|\r_a\|^5}\r_a.\label{eqn:a2}
\end{align}
The nonlinear relative orbital dynamics is locally weakly observable at $\x=[\r^T,\dot\r^T]$ if
%\begin{list}{}{\setlength{\leftmargin}{1cm}\setlength{\itemsep}{0mm}\setlength{\parsep}{0mm}\setlength{\topsep}{0mm}\setlength{\parskip}{0mm}\setlength{\labelwidth}{2cm}}
%\item[(i)] when $\r\times\dot\r =0$, $\r\times\mathbf{a}_1\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_1)\neq 0$,
%\item[(ii)] when $\r\times\dot\r \neq 0$, $\r\times\mathbf{a}_2\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_2)\neq 0$.
%\end{list}
\begin{alignat}{2}
(i)&\; &\text{when $\r\times\dot\r =0$, $\r\times\mathbf{v}_{ref}\neq0$, $\r\times\mathbf{a}_1\neq 0$,  and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_1)\neq 0$},\label{eqn:cond1}\\
(ii)& &\text{when $\r\times\dot\r \neq 0$, $\r\times\mathbf{v}_{ref}\neq0$, $\r\times\mathbf{a}_2\neq 0$ and $\r^T(\mathbf{v}_{ref}\times \mathbf{a}_2)\neq 0$.}\label{eqn:cond2}
\end{alignat}
\end{prop}

@@

We wish to show that there is no non-zero vector $\mathbf{c}$ satisfying \refeqn{Oc}. At the first three rows of \refeqn{Oc}, we have
\begin{align*}
\mathcal{O}_{00} \mathbf{c}_1 = \frac{1}{\|\r\|}(I-\y\y^T) \mathbf{c}_1 = 0.
\end{align*}
The matrix $\mathcal{O}_{00}$ has one-dimensional null space spanned by $\y$. Therefore, without loss of generality, we can choose $\mathbf{c}_1=0$ or $\mathbf{c}_1=\r$. Suppose $\mathbf{c}_1=\r$ for the subsequent development.

@@

The third condition of \refeqn{cond1} implies that three vectors $\r$, $\mathbf{v}_{rel}$ and $\mathbf{a}_1$ do not belong to the same plane. The third condition of \refeqn{cond2} has similar structure as well. This condition can be easily violated if the relative motion is planar, i.e., when $z(t)\equiv 0$ for all $t$.
